Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-06-22 paper.tex
corrections finished

\C{In Fig.~\ref{fig15}, we show flux predictions by an analytical CR model
  \cite{2010MNRAS.409..449P} for CR induced gamma-rays and compare those to the
  extended \Fermi-LAT flux upper limits that assume King profiles which trace
  the X-ray emitting gas.}  For consistency, we use beta-model fits to ROSAT
data of X-ray bright clusters for our gamma-ray flux estimates as an input
\cite{2002ApJ...567..716R}). We derive the central gas densities for each
cluster by following the approach outlined in \cite{1999ApJ...517..627M} which
assumes that the ICM is isothermal. The central gas density is found by
combining X-ray temperature with X-ray luminosity, X-ray detection radius,
density core radius and density slope for each cluster. These predictions are
superior to those that use luminosity-mass scaling relation to infer the
expected gamma-ray flux---a method used by reference \cite{2010ApJ...717L..71A}
to compare to \Fermi limits.  In this analytic model, we employ the fact that
CRs accelerated at structure formation shocks and transported over cosmic
time. This leads to a CR distribution with a universal spectral form and an
almost universal spatial distribution. Depending on whether or not we account
for the bias of ``artificial galaxies'' in cosmological simulations, we derive
an optimistic or conservative limit of the expected gamma-ray emission from
decaying pions (bracketing our ignorance of the contribution of cluster galaxies
to the overall gamma-ray flux from a cluster). Note that both models assume the
maximally allowed acceleration efficiency for strong shocks that transfers 50\%
of the shock-dissipated energy (kinetic energy corrected for adiabatic
compressional heating at the shock) to CRs. These efficiencies drop quickly for
weaker shocks \cite{2007A&A...473...41E} which dominate the gravitational energy
dissipation inside galaxy clusters \cite{2006MNRAS.367..113P}. These
efficiencies will have to be corrected down if this value is not realized at
structure formation shocks. Secondly, these simulations neglect active CR
transport such as streaming and diffusion relative to the gas, i.e. we assume
that advective transport is the dominant one and CRs are tightly coupled to the
gas via flux-frozen magnetic fields tangled on sufficiently small
scales. Evidence from the statistics of radio halos suggests that this might not
be the case for relaxed clusters, arguing that CR streaming in those objects
could be responsible for a decrease in the central CR number density. This
effect would also reduce the expected gamma-ray flux by up to a factor of five
\cite{2011A&A...527A..99E}. We note that the tightest \Fermi limits are obtained
in the 1-10 GeV regime due to the highest sensitivity of \Fermi-LAT there. While
the effective area of LAT increases up to these energies, the typical photon
spectra decrease as a function of energy: combining those effects selects this
energy range to be most sensitive.  In this energy band, the \Fermi limits come
close to the flux predictions of Virgo, Coma and Norma. Using integrated flux
limits from \Fermi, we arrive at very similar results.  We emphasize that by
using our analytic modeling, the slight discrepancy of the scaling
relation-based prediction with the \Fermi limit on Norma
\cite{2010ApJ...717L..71A} is resolved: Norma simply seems to be less bright
than an average cluster of the same mass as Norma. The insensitive limit on
Perseus \C{is due to the bright central source NGC1275 with an AGN in the center
  \cite{2010ATel.2916....1M} making it hard to determine a comparable limit as
  obtained for the other clusters. In addition, there are recent gamma-ray
  observations of the head-tail radio galaxy IC310 in the cluster outskirts
  \cite{2010ApJ...723L.207A,2010A&A...519L...6N}.}

\C{A quantity that is of great scientific interest is the CR pressure
  ($P_\CR$) relative to the thermal pressure ($P_\rmn{th}$) as it
  directly bias galaxy cluster observables and the resulting
  cosmological constraints, such as the hydrostatic mass bias or the
  Sunyaev-Zel’dovich signal measured by Planck, South Pole Telescope,
  or Atacama Cosmology Telescope. In the following we estimate the
  mass scaling of the relative CR pressure, $X_\CR =
  P_\CR/P_\rmn{th}$, using the CR formalism outlined in
  \cite{2007A&A...473...41E,2010MNRAS.409..449P}. We derive the CR
  pressure from the CR distribution function $f = C\,p^{-\alpha}$,
  where $C$ is the normalization of each power-law, $p=P/m_\rmn{p} c$
  is the normalized CR proton momentum, and $\alpha$ is the CR
  spectral index. The pressure is given by $P_\CR \propto
  C=\tilde{C} \,\rho_\rmn{gas}$, where $\tilde{C}\propto M^{0.44}$ for
  clusters with a mass $\mvir\gtrsim 10^{14}\,\msun$
  \cite{2010MNRAS.409..449P}. We determine the mass scaling of the gas
  density through its relation to the enclosed gas fraction
  $f_\rmn{gas}$, and find that $\rho_\rmn{gas}\propto f_\rmn{gas}
  \propto M^{0.135}$ \cite{2009ApJ...693.1142S}. The thermal pressure
  is derived from the temperature in virial equilibrium, and scales as
  $P_\rmn{th}\propto k_\B T \propto G M/R \propto M^{0.67}$. Finally
  we can derive the mass scaling of the volume weighted relative CR
  pressure from
\begin{equation}
 <X_\CR> = \frac{<P_\CR>}{<P_\rmn{th}>} \propto \frac{<\rho_\rmn{gas}\,
   \tilde{C}>}{<k_\B T>} \propto M^{-0.1}\,,
\end{equation}
which is roughly a constant, i.e. almost independent on halo
mass. This suggests that CRs are accelerated at the strongest shocks,
in the early universe or between voids and pre-collapsed IGM, and then
experience a similar transport history in the form of adiabatic
compression or similarly weaker shocks. Since the CRs that build up
the bulk of the CR pressure also give rise to gamma-ray emission
through the production of decaying pions, we use gamma-ray upper
limits from \Fermi-LAT to constrain $X_\CR$. In Fig.~\ref{fig:XCR} we
show the volume weighted relative CR pressure for 32 clusters, for
which LAT upper limits that assume King profiles are derived. We adopt
a constant $<X_\CR>\approx 0.02$ derived from a representative
simulated cluster \cite{2008MNRAS.385.1211P} and find that for most of
the clusters in the sample we can constrain $X_\CR < (0.1-0.2)$ using
the extended gamma-ray upper limits, where the most constraining
clusters are Norma and Coma with $X_\CR < 0.07$. Most of the 32
clusters have $X_\CR$ upper limits within a factor few from the best
constraining clusters, which suggests that stacking of clusters will
be able to improve these limits. This might even be improved further
using a scaled stacking with our model as a prior. In a recent work
\cite{2010ApJ...717L..71A}, where the 1.5 year \Fermi data was used to
derive limits on $X_\CR$, they found that the most massive clusters
are on average more constraining, while we find better limits on the
low mass clusters. This difference is explained by the different
priors on $P_\CR$; they assumed a mass independent CR pressure derived
from a single CR power-law distribution function, while we adopt a
prior inferred from cosmological simulations that dynamically trace the
CRs during cluster assembly and allow for a varying spectral index
where $P_\CR\sim M^{0.44}$. Furthermore, the gamma-ray flux
predictions that we use in this work to constrain $X_\CR$ assumes the
true gas density distribution for each cluster \footnote{Derived from
  the parameters in the extended HIFLUGCS catalogue.} that accounts
for cluster-to-cluster scatter, in contrast to the mass-to-luminosity
scaling relations that was used in previous work.}

@@

@@

In this paper, we study the possibility for detecting gamma-ray
emission in galaxy clusters. We consider benchmark (BM) and
leptophilic (LP) models of supersymmetric dark matter (DM) as well as
cosmic ray (CR) induced pion decay which is thought to dominate the
astrophysical signal from clusters. Once supersymmetric cold dark
matter decouples from the expanding universe it streams and erases any
potentially present smaller scales. When gravitational instabilities
causes halos to collapse, this free-streaming scale imprints as a
characteristic minimum halo mass $M_\mathrm{min}$ into the hierarchy
of structure formation and should still survive until the present time
as substructure within larger halos. It turns out that the boost due
to substructures dominates over the smooth component for halo masses
$\mvir>10^3 \msun$ and contributes a constant luminosity for each
decade in halo mass. The ratio of halo-to-minimum subhalo mass
$\mvir/M_\mathrm{min}$ is maximized for the largest systems that had
time to collapse and virialise since the Big Bang -- galaxy
clusters. The corresponding substructure boost can reach values
exceeding $10^3$ which should make nearby clusters the brightest DM
annihilation sources after the MW center and hypothetical very
close-by DM substructures.  Since the mass density of substructures
peaks at radii close to $\rvir$, the contribution to the annihilation
emission from substructures is also dominated from these outer cluster
regions.  This implies an almost flat surface brightness profile of
the annihilation emission which makes it necessary to have the entire
cluster in the field-of-view to take advantage of the total
substructure boost. This property makes it difficult to detect DM
annihilation emission without an enhancement of the particle physics
cross section over its standard value of $\sigma v\sim 3\times
10^{-26} ~\mathrm{cm}^3~\mathrm{s}^{-1}$ with imaging air Cherenkov
telescopes due to the spatially extended boost from substructures of
angular extent $\sim 0.5-1\deg$ (as independently found by reference
\cite{2011arXiv1104.3530S}). In contrast, the sensitivity of Cherenkov
telescopes drops linearly with radius outside the point spread
function $\sim 0.1\deg$, unless new background subtraction schemes can
be found, e.g., for the survey mode of CTA.

We identify Fornax to be the best cluster/group candidate for indirect
DM studies as it is expected to emit the brightest annihilation flux
and more importantly, has a comparably low CR induced gamma-ray flux.
\del{This in principle should allow to detect or constrain some BM
  models that have been chosen for the large gamma-ray flux yield with
  three-year \Fermi data (or constrain the substructure boost
  factor).}  Other nearby bright sources of the extended HIFLUGCS
sample of nearby bright X-ray cluster/groups are M49, Centaurus, and
NGC~4636.  The non-detection of gamma-ray emission by \Fermi in a
total of 26 clusters/groups (the mentioned clusters as well as Virgo,
Coma, Hydra, and Norma among others) considerably constrains LP models
for DM annihilation which were introduced to explain the increasing
positron-to-lepton ratio beyond 10 GeV. \C{Assuming that the minimum
  DM substructure mass extends down to an Earth mass and that the DM
  annihilation flux is dominated by small scale substructure, we limit
  the saturated Sommerfeld enhancement factor in M49 and Fornax to
  about 14 and 22, respectively. This corresponds to a value in the
  Milky Way (MW) of less than 8 and 13, respectively, due to the larger Sommerfeld
  boost factor in smaller mass halos which have a lower velocity
  dispersion that enhances the particle cross section (while assuming
  universality of the DM model).} This would rule out LP models in
their current form based on the non-observation of gamma-rays in any
of the fore-mentioned clusters/groups. Alternatively, assuming the LP
models to be correct, we limit the minimum substructure mass to
$>10~\msun$ in M49.

In this work, we thoroughly study the spectral emission
characteristics of the DM emission. In general there are different
radiation mechanism that emit in the gamma-ray regime, namely (i)
continuum emission following the hadronization of the annihilating
neutralinos in BM models which leads to the production of charged and
neutral mesons which decay finally into electrons, neutrinos, and
gammas, (ii) inverse Compton (IC) emission by the leptons in the final
state which up-scatter radiation fields from the cosmic microwave
background (CMB), dust, and starlight, and (iii) final state radiation
or internal bremsstrahlung which dominates the highest energy emission
close to the rest mass of the self-annihilating neutralinos (if
present in the model under consideration). While this final state
radiation is sometimes not distinguished from the continuum emission
contribution, we keep it separate as it dominates the very high-energy
tail of the gamma-ray spectrum in LP models while there the continuum
emission is negligibly low.

For the first time, we compute the IC emission component not only from
the upscattered CMB photons but also from a realistically modeled
photon field due to dust and starlight extending from infra-red to
optical wavelengths. To this end, we construct a rather simple
analytic model for the spatial and spectral distribution of dust and
starlight in clusters (which can also be used for the annihilation
emission in galaxies after adopting a different spatial profile).  We
find that in BM models, the continuum emission dominates over the IC
emission of any available radiation field. In the LP models, where the
continuum emission is low, the IC up-scattering of the CMB dominates
below 60 GeV. \C{For larger energies, the final state radiation
  dominates the emission up to the cutoff at $m_\chi c^2 \sim
  1$~TeV. The IC up-scattering of the dust and starlight is always
  subdominant compared to the upscattered CMB (below 60~GeV) and the
  final state radiation (above 60~GeV).}  \del{For large clusters, the
  final state radiation then takes over for energies up to the cutoff
  at $m_\chi c^2 \sim 1$~TeV. In small clusters, there is a window
  around $E_\gamma \sim 100~\mathrm{GeV}$ where the IC up-scattering
  of the dust emission dominates over the CMB IC emission and the
  final state radiation. The IC emission of starlight remains always
  subdominant.}  We stress, that the inclusion of the dust and
starlight radiation fields are negligible in the cooling function for
the leptons.  The main reason why the IC emission from starlight and
dust remains subdominant is the small overlap of this component with
the peripherally peaked substructure mass density profiles. If for
some reason, DM has less substructure, the relative importance of this
IC component increases considerably.

We substantially improve the modelling of the expected gamma-ray
signal from pion-decay resulting from hadronic CR interactions with
gas protons over previous work \citep{2010ApJ...717L..71A}. We employ
the analytic model of the CR spectrum and spatial distribution,
recently developed by \citet{2010MNRAS.409..449P} that is based on
cosmological hydrodynamical simulations of galaxy clusters that
self-consistently follow the evolution of CRs in cosmic structure
formation. Adopting the model for all clusters in the HIFLUGCS sample
of flux-limited X-ray clusters, we compare the expected gamma-ray
emission to that inferred from using $L_\gamma$-cluster mass scaling
relation. We note that the predictions of our ``optimistic'' model are
perfectly in agreement with \Fermi upper limits for all clusters.  The
hint of a tension that the scaling relation modeling showed for Norma
\citep{2010ApJ...717L..71A} could not be confirmed due to the lower
gas density of this cluster relative to the average.  As expected, the
analytical CR model accounts for the ``scatter'' to the scaling
relation and biases the gamma-ray flux high for prominent cool core
clusters by up to a factor of 6 relative to the expectation from the
scaling relation. This is due to the high target gas densities and
associated CR densities in those cool cores that shape the spatial
emission characteristic to be very similar to that observed in thermal
X-rays. We caution that these predictions only apply for the case of
negligible active CR transport such as CR streaming and diffusion
relative to the gas. Gamma-ray fluxes might be considerably smaller in
those cluster that do not show an extended diffuse radio(-mini) halo
that might point to a centrally concentrated CR distribution
\citep{2011A&A...527A..99E}. The CR spectrum $E^2 dN/dE$ shows the
characteristic pion bump at energies around 1~GeV. It is expected to
dominate the DM annihilation signal of BM models for all
clusters/groups but Fornax. \C{We conclude that due to the large
  spatial extension of the brightest clusters it is not possible for
  \Fermi to rule out the LP model or to constrain the BM DM models
  without improved modelling. However, we can put tight constraints on
  the boost from substructures and Sommerfeld enhancement using the LP
  model. Hence, clusters are a valuable complement to direct and
  accelerator searches for the nature of DM.}\del{We conclude that it
  will be possible to rule out the LP model with six years of \Fermi
  data and start to constrain some BM DM models. Hence this should
  nicely complement direct and accelerator searches for the nature of
  DM.}

>>>>

@@

\C{Figure~\ref{fig15} shows flux predictions by an analytical CR model
  \cite{2010MNRAS.409..449P} for CR induced gamma-rays and compare those to the
  extended \Fermi-LAT flux upper limits that assume King profiles which trace
  the X-ray emitting gas.}  For consistency, we use beta-model fits to ROSAT
data of X-ray bright clusters for our gamma-ray flux estimates as an input
\cite{2002ApJ...567..716R}). We derive the central gas densities for each
cluster by following the approach outlined in \cite{1999ApJ...517..627M} which
assumes that the ICM is isothermal. The central gas density is found by
combining X-ray temperature with X-ray luminosity, X-ray detection radius,
density core radius and density slope for each cluster. These predictions are
superior to those that use luminosity-mass scaling relation to infer the
expected gamma-ray flux---a method used by reference \cite{2010ApJ...717L..71A}
to compare to \Fermi limits.  In this CR analytic model, we model
CR-acceleration at structure formation shocks and their transport over cosmic
time. This leads to a CR distribution with a universal spectral form and an
almost universal spatial distribution. Depending on whether or not we account
for the bias of ``artificial galaxies'' in cosmological simulations, we derive
an optimistic or conservative limit of the expected gamma-ray emission from
decaying pions (bracketing our ignorance of the contribution of cluster galaxies
to the overall gamma-ray flux from a cluster).

\C{We highlight two main uncertainties of the models, namely the acceleration
  efficiency at formation shocks and CR transport parameters. First, we note}
that both models assume the maximally allowed acceleration efficiency for {\em
  strong} shocks that transfers 50\% of the shock-dissipated energy (kinetic
energy corrected for adiabatic compressional heating at the shock) to CRs. These
efficiencies drop quickly for weaker shocks \cite{2007A&A...473...41E} which
dominate the gravitational energy dissipation inside galaxy clusters
\cite{2006MNRAS.367..113P}. These efficiencies will have to be corrected down if
this value is not realized at structure formation shocks. Second, these
simulations \C{(and by extension the analytic model)} neglect active CR
transport such as streaming and diffusion relative to the gas, i.e., we assume
that advective transport dominates and CRs are tightly coupled to the gas via
flux-frozen magnetic fields tangled on sufficiently small scales. Evidence from
the statistics of radio halos suggests that this might not be the case for
relaxed clusters, arguing that CR streaming \C{(and diffusion)} in those objects
could be responsible for a decrease in the central CR number density. This
effect would also reduce the expected gamma-ray flux by up to a factor of five
\cite{2011A&A...527A..99E}.

We note that the tightest \Fermi limits are obtained in the 1-10 GeV regime due
to the highest sensitivity of \Fermi-LAT there. While the effective area of LAT
increases up to these energies, the typical photon spectra decrease as a
function of energy: combining those effects selects this energy range to be most
sensitive.  In this energy band, the \Fermi limits come close to the flux
predictions of Virgo, Coma and Norma. Using integrated flux limits from \Fermi,
we arrive at very similar results.  We emphasize that by using our analytic
modeling, the slight discrepancy of the scaling relation-based prediction with
the \Fermi limit on Norma \cite{2010ApJ...717L..71A} is resolved: Norma simply
seems to be less bright than an average cluster of the same mass as Norma. The
insensitive limit on Perseus \C{is due to the bright central source NGC1275 with
  an AGN in the center \cite{2010ATel.2916....1M} making it hard to determine a
  comparable limit as obtained for the other clusters. In addition, there are
  recent gamma-ray observations of the head-tail radio galaxy IC310 in the
  cluster outskirts \cite{2010ApJ...723L.207A,2010A&A...519L...6N}.}

\C{A quantity that is of great scientific interest is the CR pressure ($P_\CR$)
  relative to the thermal pressure ($P_\rmn{th}$) as it may bias galaxy cluster
  observables, e.g., the Sunyaev-Zel’dovich signal or hydrostatic mass
  estimates, and the resulting cosmological constraints if the analyses do not
  marginalize over a potential pressure contribution from CRs. Before we turn to
  the newly derived constraints on the relative CR pressure, $X_\CR =
  P_\CR/P_\rmn{th}$, we study the Bayesian prior that different models (a
  simulation based and a simple analytic model) impose on the scaling of $X_\CR$
  with cluster mass.}

\C{In the simulation based analytic CR model \cite{2010MNRAS.409..449P}, the CR
  pressure derives from the CR distribution function, $f\dd p =
  C\,p^{-\alpha}\dd p$, where $C$ is the normalization, $p=P/m_\rmn{p} c$ is the
  normalized CR proton momentum, and $\alpha$ is the CR spectral index. The
  pressure is given by $P_\CR \propto C\propto \tilde{C} \,\rho_\rmn{gas}$,
  where the volume weighted, dimensionless normalization,
  $\bra\tilde{C}\ket\propto \mvir^{0.44}$ for clusters with a mass $\mvir\gtrsim
  10^{14}\,\msun$ \cite{2010MNRAS.409..449P}. The thermal pressure is derived
  from the temperature in virial equilibrium, and scales as $P_\rmn{th}= n k_\B
  T \propto G \mvir/R_{200} \propto \mvir^{2/3}$. Hence, the volume weighted
  relative CR pressure scales as
  \begin{equation}
    \bra X_{\CR,\,\rmn{sim}}\ket = \frac{\bra P_\CR\ket}{\bra P_\rmn{th}\ket} 
    \propto \frac{\bra\tilde{C}\ket}{\bra k_\B T\ket} \propto M^{-0.22}\,,
  \end{equation}
  which depends only weakly on cluster virial mass. This suggests that
  CRs are accelerated at the strongest formation shocks, i.e. during
  proto-cluster formation at high$-z$ or in between voids and
  pre-collapsed IGM. These CRs experience then a similar transport
  history in the form of adiabatic compression or encounter shocks of
  similar (weak) strength.}

\C{In contrast, the analytic (isobaric) CR model \cite{2004A&A...413...17}, that
  was used by the \Fermi collaboration to derive CR pressure constraints
  \cite{2010ApJ...717L..71A}, assumes $P_\CR = X_\CR P_\rmn{th}$ and constrains
  $X_\CR$ by assuming no intrinsic cluster mass dependence of the CR pressure.
  While apparently very natural, this implies a strong cluster mass dependence
  of $X_\CR$ as the following estimate shows:
  \begin{equation}
    \bra X_{\CR,\,\rmn{iso}}\ket = \frac{\bra P_\CR\ket}{\bra P_\rmn{th}\ket} 
    \propto \frac{C}{\bra n k_\B T\ket} \propto M^{-0.8}\,.
  \end{equation}
  Here, we determine the mass scaling of the average gas density
  through its relation to the enclosed gas fraction, $f_\rmn{gas}$, and find that
  $n_\rmn{gas}\propto f_\rmn{gas} \propto M^{0.135}$
  \cite{2009ApJ...693.1142S}.}

\C{Figure~\ref{fig:XCR} shows the volume weighted relative CR pressure for 32
  clusters, for which extended \Fermi-LAT upper limits are derived (that assume
  King profiles). We adopt a constant $\bra X_\CR\ket\approx 0.02$ derived from
  the average of the large clusters in our simulation sample
  \cite{2008MNRAS.385.1211P, 2010MNRAS.409..449P}. We can constrain $X_\CR <
  (0.1-0.2)$ using the extended gamma-ray upper limits on most of the clusters
  in the \Fermi sample. The most constraining clusters are Norma and Coma with
  $X_\CR < 0.07$. Most of the 32 clusters have limits on $X_\CR$ within a factor
  few from the best constraining clusters suggesting that stacking of clusters
  will be able to improve these limits.  These limits may improved upon by using
  a stacking approach that adopts our simulation-based analytic model as a
  prior.}

\C{In a recent work by the \Fermi collaboration \cite{2010ApJ...717L..71A},
  where the 1.5 year \Fermi data was used to derive limits on $X_\CR$, they
  found that the most massive clusters are on average more constraining yielding
  the best constraints of the order of 5\% (by adopting, however, point-like
  upper limits). In contrast, we find better limits on the low mass
  clusters. This difference is explained by the different priors on $P_\CR$.  We
  adopt a prior inferred from cosmological simulations that dynamically trace
  the CRs during cluster assembly suggesting that adiabatic transport is
  dominant in shaping the CR population in clusters and implying a weak mass
  dependence of $X_\CR$. Their analysis assumes a CR pressure that is
  independent of cluster mass, i.e., constraints on $X_\CR$ inherit the inverse
  mass dependence of the thermal pressure, $X_\CR\propto 1/P_\rmn{th}\propto
  M^{-0.8}$. While it is not clear which mass scaling of $X_\CR$ is realized in
  Nature, these considerations suggest that it will be critical for future work
  to account for the Bayesian prior in deriving limits on their pressure.}

 %%%%%
\del{ In the following we
  estimate the mass scaling of the relative CR pressure, $X_\CR =
  P_\CR/P_\rmn{th}$, using the CR formalism outlined in
  \cite{2007A&A...473...41E,2010MNRAS.409..449P}. We derive the CR pressure from
  the CR distribution function $f = C\,p^{-\alpha}$, where $C$ is the
  normalization of each power-law, $p=P/m_\rmn{p} c$ is the normalized CR proton
  momentum, and $\alpha$ is the CR spectral index. The pressure is given by
  $P_\CR \propto C=\tilde{C} \,\rho_\rmn{gas}$, where $\tilde{C}\propto
  M^{0.44}$ for clusters with a mass $\mvir\gtrsim 10^{14}\,\msun$
  \cite{2010MNRAS.409..449P}. We determine the mass scaling of the gas density
  through its relation to the enclosed gas fraction $f_\rmn{gas}$, and find that
  $\rho_\rmn{gas}\propto f_\rmn{gas} \propto M^{0.135}$
  \cite{2009ApJ...693.1142S}. The thermal pressure is derived from the
  temperature in virial equilibrium, and scales as $P_\rmn{th}\propto k_\B T
  \propto G M/R \propto M^{0.67}$. Finally we can derive the mass scaling of the
  volume weighted relative CR pressure from
  \begin{equation}
    <X_\CR> = \frac{<P_\CR>}{<P_\rmn{th}>} \propto \frac{<\rho_\rmn{gas}\,
      \tilde{C}>}{<k_\B T>} \propto M^{-0.1}\,,
  \end{equation}
  which is roughly a constant, i.e. almost independent on halo
  mass. This suggests that CRs are accelerated at the strongest
  shocks, in the early universe or between voids and pre-collapsed
  IGM, and then experience a similar transport history in the form of
  adiabatic compression or similarly weaker shocks. Since the CRs that
  build up the bulk of the CR pressure also give rise to gamma-ray
  emission through the production of decaying pions, we use gamma-ray
  upper limits from \Fermi-LAT to constrain $X_\CR$. In
  Fig.~\ref{fig:XCR} we show the volume weighted relative CR pressure
  for 32 clusters, for which LAT upper limits that assume King
  profiles are derived. We adopt a constant $<X_\CR>\approx 0.02$
  derived from a representative simulated cluster
  \cite{2008MNRAS.385.1211P} and find that for most of the clusters in
  the sample we can constrain $X_\CR < (0.1-0.2)$ using the extended
  gamma-ray upper limits, where the most constraining clusters are
  Norma and Coma with $X_\CR < 0.07$. Most of the 32 clusters have
  $X_\CR$ upper limits within a factor few from the best constraining
  clusters, which suggests that stacking of clusters will be able to
  improve these limits. This might even be improved further using a
  scaled stacking with our model as a prior. In a recent work
  \cite{2010ApJ...717L..71A}, where the 1.5 year \Fermi data was used
  to derive limits on $X_\CR$, they found that the most massive
  clusters are on average more constraining, while we find better
  limits on the low mass clusters. This difference is explained by the
  different priors on $P_\CR$; they assumed a mass independent CR
  pressure derived from a single CR power-law distribution function,
  while we adopt a prior inferred from cosmological simulations that
  dynamically trace the CRs during cluster assembly and allow for a
  varying spectral index where $P_\CR\sim M^{0.44}$. Furthermore, the
  gamma-ray flux predictions that we use in this work to constrain
  $X_\CR$ assumes the true gas density distribution for each cluster
  \footnote{Derived from the parameters in the extended HIFLUGCS
    catalogue.} that accounts for cluster-to-cluster scatter, in
  contrast to the mass-to-luminosity scaling relations that was used
  in previous work.}

@@

In this paper, we study the possibility for detecting gamma-ray
emission in galaxy clusters. We consider benchmark (BM) and
leptophilic (LP) models of supersymmetric dark matter (DM) as well as
cosmic ray (CR) induced pion decay which is thought to dominate the
astrophysical signal from clusters.

{\em Dark matter annihilation.} Once supersymmetric cold dark matter decouples
from the expanding universe it streams and erases any potentially present
smaller scales. When gravitational instability causes halos to collapse, this
free-streaming scale imprints as a characteristic minimum halo mass
$M_\mathrm{min}$ into the hierarchy of structure formation and should still
survive until the present time as substructure within larger halos. We show that
the boost due to substructures dominates over the smooth component for halo
masses $\mvir>10^3 \msun$ and contributes a constant luminosity for each decade
in halo mass. The ratio of halo-to-minimum subhalo mass $\mvir/M_\mathrm{min}$
is maximized for the largest virialized systems in the Universe -- galaxy
clusters. The corresponding substructure boost can reach values exceeding $10^3$
which should make nearby clusters the brightest DM annihilation sources after
the MW center and hypothetical very close-by DM substructures. \C{We stress that
  this conclusion relies on two hypotheses that are still open problems in
  particle physics and numerical cosmology. We assume (i) the existence of cold
  DM, i.e., the density fluctuation power spectrum extends down to the DM free
  streaming scale of about an Earth mass, and (ii) assume that structure
  formation at scales smaller than currently resolved ones ($\sim 10^5\,\msun$)
  proceeds scale-invariant, i.e., there is no non-linear mode coupling between
  different scales that could potentially erase structure smaller than a
  characteristic one as the linear power spectrum approaches $\Delta^2(k)
  \propto k^3 P(k)=\rmn{const.}$ and structure on different scales forms not any
  more hierarchically, but collapses at the same time.}

Since the mass density of substructures peaks at radii close to
$\rvir$, the contribution to the annihilation emission from
substructures is also dominated from these outer cluster regions.
This implies an almost flat surface brightness profile of the
annihilation emission which makes it necessary to have the entire
cluster in the field-of-view to take advantage of the total
substructure boost. This property makes it difficult to detect DM
annihilation emission without an enhancement of the particle physics
cross section over its standard value of $\sigma v\sim 3\times
10^{-26} ~\mathrm{cm}^3~\mathrm{s}^{-1}$ with imaging air Cherenkov
telescopes due to the spatially extended boost from substructures of
angular extent $\sim 0.5-1\deg$ (as independently found by reference
\cite{2011arXiv1104.3530S}). In contrast, the sensitivity of Cherenkov
telescopes drops linearly with radius outside the point spread
function $\sim 0.1\deg$, unless new background subtraction schemes can
be found, e.g., for the survey mode of CTA.

In this work, we thoroughly study the spectral emission
characteristics of the DM emission. In general there are different
radiation mechanism that emit in the gamma-ray regime, namely (i)
continuum emission following the hadronization of the annihilating
neutralinos in BM models which leads to the production of charged and
neutral mesons which decay finally into electrons, neutrinos, and
gammas, (ii) inverse Compton (IC) emission by the leptons in the final
state which up-scatter radiation fields from the cosmic microwave
background (CMB), dust, and starlight, and (iii) final state radiation
or internal bremsstrahlung which dominates the highest energy emission
close to the rest mass of the self-annihilating neutralinos (if
present in the model under consideration).\footnote{While this final
  state radiation is sometimes not distinguished from the continuum
  emission contribution, we keep it separate as it dominates the very
  high-energy tail of the gamma-ray spectrum in LP models while there
  the continuum emission is negligibly low.}

For the first time, we compute the IC emission component not only from
the upscattered CMB photons but also from a realistically modeled
photon field due to dust and starlight extending from infra-red to
optical wavelengths. To this end, we construct a rather simple
analytic model for the spatial and spectral distribution of dust and
starlight in clusters (which can also be used for the annihilation
emission in galaxies after adopting a different spatial profile).  We
find that in BM models, the continuum emission dominates over the IC
emission of any available radiation field. In the LP models, where the
continuum emission is low, the IC up-scattering of the CMB dominates
below 60 GeV. \C{For larger energies, the final state radiation
  dominates the emission up to the cutoff at $m_\chi c^2 \sim
  1$~TeV. The IC up-scattering of the dust and starlight is always
  subdominant compared to the upscattered CMB (below 60~GeV) and the
  final state radiation (above 60~GeV).}  We stress, that the
inclusion of the dust and starlight radiation fields are negligible in
the cooling function for the leptons.  The main reason why the IC
emission from starlight and dust remains subdominant is the small
overlap of this component with the peripherally peaked substructure
mass density profiles. If for some reason, DM has less substructure,
the relative importance of this IC component increases considerably.

\C{We identify Virgo and Fornax to be the best cluster/group
  candidates for indirect DM studies as they are expected to emit the
  brightest annihilation flux.\footnote{\C{We excluded Virgo from our
    further analysis due to its large angular extent that needs a
    separate treatment.}}  More importantly, Fornax has a comparably
  low CR induced gamma-ray flux which may enable an indirect DM
  detection or a particularly tight limit on DM properties}.  Other
nearby bright sources of the extended HIFLUGCS sample of nearby bright
X-ray cluster/groups are \C{Ophiuchus, Coma, and Centaurus}.

The non-detection of gamma-ray emission by \Fermi in a total of 26
clusters/groups considerably constrains LP models for DM annihilation
which were introduced to explain the increasing positron-to-lepton
ratio beyond 10 GeV. \C{Assuming that the minimum DM substructure mass
  extends down to an Earth mass and that the DM annihilation flux is
  dominated by small scale substructure, we limit the saturated
  Sommerfeld enhancement factor in M49 (that has a particularly tight
  upper limit) and Fornax to about 14 and 22, respectively. This
  corresponds to a value in the Milky Way (MW) of less than 8 and 13,
  respectively, due to the larger Sommerfeld boost factor in smaller
  mass halos which have a lower velocity dispersion that enhances the
  particle cross section (while assuming universality of the DM
  model).} This would rule out LP models in their current form based
on the non-observation of gamma-rays in any of the fore-mentioned
clusters/groups. Alternatively, assuming the LP models to be correct,
we limit the minimum substructure mass to $>10~\msun$ in M49.

{\em Cosmic ray induced emission.} We substantially improve the
modelling of the expected gamma-ray signal from pion-decay resulting
from hadronic CR interactions with gas protons over previous work
\citep{2010ApJ...717L..71A}. We employ the analytic model of the CR
spectrum and spatial distribution that is based on cosmological
hydrodynamical simulations of galaxy clusters that self-consistently
follow the evolution of CRs in cosmic structure formation
\citet{2010MNRAS.409..449P}. Adopting the model for all clusters in
the HIFLUGCS sample of flux-limited X-ray clusters, we compare the
expected gamma-ray emission to that inferred from using
$L_\gamma$-cluster mass scaling relation. We note that the predictions
of our ``optimistic'' model are perfectly in agreement with \Fermi
upper limits for all clusters.  The hint of a tension that the scaling
relation modeling showed for Norma \citep{2010ApJ...717L..71A} could
not be confirmed due to the lower gas density of this cluster relative
to the average.

As expected, the analytical CR model accounts for the ``scatter'' to
the scaling relation and biases the gamma-ray flux high for prominent
cool core clusters by up to a factor of 6 relative to the expectation
from the scaling relation. This is due to the high target gas
densities and associated CR densities in those cool cores that shape
the spatial emission characteristic to be very similar to that
observed in thermal X-rays. We caution that these predictions only
apply for the case of negligible active CR transport such as CR
streaming and diffusion relative to the gas. Gamma-ray fluxes might be
considerably smaller in those cluster that do not show an extended
diffuse radio(-mini) halo that might point to a centrally concentrated
CR distribution \citep{2011A&A...527A..99E}. The CR spectrum $E^2
dN/dE$ shows the characteristic pion bump at energies around 1~GeV. It
is expected to dominate the DM annihilation signal of BM models for
all clusters/groups but Fornax.

\C{Combining extended \Fermi upper limits and our model flux predictions, we
  limit the relative CR pressure, $X_\CR = P_\CR/P_\rmn{th}$, in 32 nearby
  bright galaxy clusters of the \Fermi sample. The best limits are found in
  Norma and Coma of the order of 7\%, with typical limits ranging between 10\%
  to 30\%. This is somewhat larger that those inferred by the \Fermi
  collaboration \citep{2010ApJ...717L..71A} and due to the differently assumed
  Bayesian prior on the CR pressure which implies a different cluster mass
  dependence of the resulting limits on $X_\CR$. }

\C{We conclude that due to the large spatial extension of the
  brightest clusters it is not possible for \Fermi to rule out the LP
  model or to constrain the BM DM models without improved
  modelling. However, we can put tight constraints on the boost from
  substructures and Sommerfeld enhancement using the LP model. Hence,
  clusters are a valuable complement to direct and accelerator
  searches for the nature of DM.}
